---
title :
Author: Maria Kondili
Date: October 2022
output:
  html_document :
    toc: true
    highlight: tango
    df_print: paged
editor_options:
  chunk_output_type: console
---

### Subject: Explore GSEA using genes from proteomics data

```{r}

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(AnnotationDbi))
suppressPackageStartupMessages(library(clusterProfiler))
suppressPackageStartupMessages(library(pathview))
suppressPackageStartupMessages(library(enrichplot))
suppressPackageStartupMessages(library(org.Hs.eg.db))
suppressPackageStartupMessages(library(DOSE))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(fgsea))
library(here)
```



```{r}

proj_dir <- here()

setwd(proj_dir)

dir.create("GSEA_Output/",showWarnings = FALSE)

```

### Read INPUT from "Stats" table of 3P5-proteomics platform

- ALL

```{r}

all_surnag_stats <- readr::read_delim("Data/All/Surnageant_All_Proteins_Secretome_Stats.txt",delim="\t",col_names = T)
all_culot_stats  <- readr::read_delim("Data/All/Culots_All_Proteins_Secretome_Stats.txt",delim="\t",col_names = T)


cat("Totally the supernatant proteins measured are : \n")
nrow(all_surnag_stats)

cat("Totally the Cellular proteins measured are : \n")
nrow(all_culot_stats)

```

### PATHWAYS  analysis

```{r}
pathw_dir="~/MSigDB_Hallmark_Pathways/"

analyse_gsea <- function(input_data, pathw_dir, source="Surnageant"){

  gsea_genes <- input_data$LGMDvsCTRL_FoldChange
  names(gsea_genes) <- input_data$Gene_name
  gsea_genes <- sort(gsea_genes, decreasing=T)

  ###
  ### Convert in ENTREZ.ID
  ###

  source("~/myScriptsBank/convert_GeneIds.R")
  entrezid          <- convert_gene_ids.hs(names(gsea_genes), from="SYMBOL", to="ENTREZID")
  names(gsea_genes) <- entrezid
  gsea_genes        <- gsea_genes[-duplicated(names(gsea_genes))]


  ###
  ### FGSEA
  ###

  c2 <- gmtPathways(paste0(pathw_dir, "C2_curated_gene_sets_Canonical_Pathways_Hs_EntrezIDs.gmt"))
  c5 <- gmtPathways(paste0(pathw_dir, "C5_ontology_gene_sets_all_2022_Hs_EntrezIDs.gmt"))
  hm <- gmtPathways(paste0(pathw_dir, "H_Hallmark_gene_sets_all_2022_Hs_EntrezIDs.gmt"))
  gseaRes.c2 <- fgsea(c2, gsea_genes)
  gseaRes.c5 <- fgsea(c5, gsea_genes)
  gseaRes.hm <- fgsea(hm, gsea_genes)

  gseaRes <- bind_rows( gseaRes.c2, gseaRes.c5,gseaRes.hm )

  top_gseaRes <- subset(gseaRes,padj < 0.1)[1:20,]

  #ns_gseaRes <-  subset(gseaRes.all,padj > 0.05)

  g<- ggplot(top_gseaRes,
             aes(reorder(pathway, NES), NES)) +
    geom_col(aes(fill= NES>0),size=1.5) +
    coord_flip() +
    labs(x="Pathway", y="Normalized Enrichment Score",
         title= paste0(source, ": All signif Pathways(padj<0.1)")) +
    theme(axis.text = element_text(size = 7)) +
    theme(axis.title = element_text(size = 12))
  print(g)

  return(gseaRes)

}
```


#### RUN function for GSEA :

> ALL :

```{r gsea_all}
gseaRes.surnag.all <- analyse_gsea(input_data=all_surnag_stats, pathw_dir, source="Surnageant")

gseaRes.cells.all  <- analyse_gsea(input_data=all_culot_stats, pathw_dir, source="Cells")

## Write in tables :
dir.create("Output",showWarnings = FALSE)

gseaRes.cells.all <- gseaRes.cells.all[order(gseaRes.cells.all$padj,decreasing = F), ]
#openxlsx::write.xlsx(gseaRes.cells.all, "GSEA_Output/GSEA_pathways_for_Culot_data_All_Proteins.xlsx")


gseaRes.surnag.all <- gseaRes.surnag.all[order(gseaRes.surnag.all$padj,decreasing = F),]
#openxlsx::write.xlsx(gseaRes.surnag.all,"GSEA_Output/GSEA_pathways_for_Surnageant_data_All_Proteins.xlsx")
```

> POI :

```{r gsea_poi}

poi_surnag_stats <- readr::read_delim("Data/POI/Surnageant_POI_Secretome_Stats.txt",delim="\t",col_names = T)

poi_culot_stats  <- readr::read_delim("Data/POI/Culots_POI_Secretome_Stats.txt",    delim="\t",col_names = T)


gseaRes.surnag.poi <- analyse_gsea(input_data = poi_surnag_stats, pathw_dir, source="Surnageant")
## ! NONE signif !

gseaRes.cells.poi  <- analyse_gsea(input_data = poi_culot_stats,  pathw_dir,  source="Cells")

#openxlsx::write.xlsx(gseaRes.cells.poi , "GSEA_Output/GSEA_pathways_Culots_POI.xlsx")

```

### Cellular-Compartments GO-terms where Culot-proteins are involved :

 >  CC @ Culot

  Cannot search for CC @ Surnageant-POI bcs NO GSEA-Pathways found enriched

```{r cc_for_culot_poi}

gseaRes.cells.poi.GOCC <- gseaRes.cells.poi[which(str_detect(gseaRes.cells.poi$pathway, "GOCC_")),]
# 310 pathways found

writexl::write_xlsx(gseaRes.cells.poi.GOCC, "GSEA_Output/Cellular_Compartments_of_Culot_Genes_GSEA_results.xlsx")

top_gseaRes.GOCC <- bind_rows(subset(gseaRes.cells.poi.GOCC,NES >  1)[1:10,],
                              subset(gseaRes.cells.poi.GOCC,NES < -1)[1:10,])

ggplot(top_gseaRes.GOCC[1:30,],
       aes(reorder(pathway, NES), NES)) +
      geom_col(aes(fill= NES>0),size=1.5) +
      coord_flip() +
      labs(x="Pathway", y="Normalized Enrichment Score",
           title="Culot:Top GSEA terms ") +
      theme(axis.text = element_text(size = 7)) +
      theme(axis.title = element_text(size = 12)  )

```

> CC @ Surnageant-ALL

```{r cc_for_surnag_all}

gseaRes.surnag.all.GOCC <- gseaRes.surnag.all[which(str_detect(gseaRes.surnag.all$pathway, "GOCC_")),]

writexl::write_xlsx(gseaRes.surnag.all.GOCC, "GSEA_Output/Cellular_Compartments_of_Surnageant_Genes_GSEA_results.xlsx")

top_gseaRes.surnag.GOCC <- bind_rows(subset(gseaRes.cells.poi.GOCC,NES >  1)[1:10,],
                              subset(gseaRes.cells.poi.GOCC,NES < -1)[1:10,])

ggplot(top_gseaRes.surnag.GOCC[1:30,],
       aes(reorder(pathway, NES), NES)) +
      geom_col(aes(fill= NES>0),size=1.5) +
      coord_flip() +
      labs(x="Pathway", y="Normalized Enrichment Score",
           title="Surnageant:Top CC-terms ") +
      theme(axis.text = element_text(size = 7)) +
      theme(axis.title = element_text(size = 12)  )

```

#### Extract genes/proteins of our data contained in each Gene-Set Enriched:

```{r}

c2 <- gmtPathways(paste0(pathw_dir, "C2_curated_gene_sets_Canonical_Pathways_Hs_EntrezIDs.gmt"))
c5 <- gmtPathways(paste0(pathw_dir, "C5_ontology_gene_sets_all_2022_Hs_EntrezIDs.gmt"))
hm <- gmtPathways(paste0(pathw_dir, "H_Hallmark_gene_sets_all_2022_Hs_EntrezIDs.gmt"))

all_pathw <- c(c2,c5,hm)

all_pathw[1:5]
```

The "all_pathw" gmtPathways contain the pathways with the genes included

We will extract the genes of our Proteomics that are present in the enriched-terms of the protein.data

### Function

```{r function_which_poi_in_pathway}

extract_genes_in_enrichPathways <- function(gseaRes, all_pathw, prot_stats_table,out_filename, write_down=FALSE) {

    source("~/myScriptsBank/convert_GeneIds.R")

    #idx_pathw_match <- which(names(all_pathw) %in% gseaRes$pathway)

    pos_pathw_match        <- match(names(all_pathw),gseaRes$pathway)
    names(pos_pathw_match) <- seq(1,length(names(all_pathw)))
    pos_pathw_match        <- pos_pathw_match[- which(is.na(pos_pathw_match))]

    genes_match <- all_pathw[as.integer(names(pos_pathw_match))]

    #length(pos_pathw_match) == length(genes_match)

    ## pathways in list with genes in symbols
    geneSymbols <- map(genes_match, ~convert_gene_ids.hs(.x, from="ENTREZID",to="SYMBOL"))

    names(geneSymbols) <- names(genes_match)

    ## find which genes of Proteomics-dataset are included in enriched pathways:
    geneSymbols_match <- map(geneSymbols,
                            ~(prot_stats_table$Gene_name[which(prot_stats_table$Gene_name %in% .x)]))
    ##>transform in Data-frame:
    genes_in_pathw_df <- data.frame("pathway"= names(geneSymbols_match))

    genesets_colaps <- purrr::map(geneSymbols_match, ~(paste(.x,collapse=",")))

    genes_in_pathw_df <- mutate(genes_in_pathw_df,
                                "my_geneset" = unlist(genesets_colaps),
                                "padj"= gseaRes$padj[pos_pathw_match],
                                "NES"=  gseaRes$NES[pos_pathw_match])

    #> many pathways without Our genes matching -> remove
    genes_in_pathw_df  <- subset(genes_in_pathw_df, my_geneset != "")

    if (write_down == TRUE) {
      writexl::write_xlsx(genes_in_pathw_df, out_filename)
      #write_delim(genes_in_pathw_df, file=out_filename,delim="\t",col_names=T)
      cat("\nFinal Table with Pathways and Genes has been saved at : \n")
      cat(out_filename)
    }

    return( genes_in_pathw_df)
}

```

> for Surnageant-Proteins :

```{r call_fun_surnag_poi}

extract_genes_in_enrichPathways(gseaRes.surnag.all, all_pathw, poi_surnag_stats,
                                out_filename = "GSEA_Output/POI_Genes_Surnageant_in_GSEA_pathways.xlsx",
                                write_down = T )

```

>> Surnageant proteins involved in: "Exosome", "Phagocytosis", "Autophagy" Pathways

```{r}

gseaRes.surnag.exo <- gseaRes.surnag.all[which(str_detect(gseaRes.surnag.all$pathway,
                                              "EXTRACELLULAR|EXOSOM|AUTOPHAG|PHAGOCYTOS")),]

which(gseaRes.surnag.exo$padj < 0.1 )
#! None Signif !

ggplot(gseaRes.surnag.exo[1:30,],
       aes(reorder(pathway, NES), NES)) +
      geom_col(aes(fill= NES>0),size=1.5) +
      coord_flip() +
      labs(x="Pathway", y="Normalized Enrichment Score",
           title="Surnageant:Top GSEA terms of Exosome ") +
      theme(axis.text = element_text(size = 7)) +
      theme(axis.title = element_text(size = 12))


## Find which Genes/Prot participate :
surnag_genes_in_exo <- extract_genes_in_enrichPathways(gseaRes.surnag.exo, all_pathw, poi_surnag_stats,
                                  out_filename = "GSEA_Output/POI_Surnageant_in_Exosomes-Phagocytosis.xlsx",
                                  write_down=T)


```

>> Surnageant Proteins involved in: "Adipocytes", "Metabolism of FATTY ACIDS"

```{r}

gseaRes.surnag.adip <- gseaRes.surnag.all[which(str_detect(gseaRes.surnag.all$pathway,
                                                           "ADIPO|FATTY_ACID|FAT_")),]

#>> ALL-proteins (non signif )
Surnag_genes_in_FattyAcidsPathw <- extract_genes_in_enrichPathways(gseaRes.surnag.adip, all_pathw, all_surnag_stats,
                                     out_filename = "GSEA_Output/All_Surnageant_in_Adipocytes-FattyAcids_Pathways.xlsx",
                                     write_down = T )

#>> POI-proteins
poi_surnag_in_fattyAcidsPathw <-extract_genes_in_enrichPathways(gseaRes.surnag.adip, all_pathw, poi_surnag_stats,
                                        out_filename = "GSEA_Output/POI_Surnageant_in_Adipocytes-FattyAcids.xlsx",
                                        write_down = T )

```

>> Surnageant Proteins in Cell.Compartment :

```{r}

extract_genes_in_enrichPathways(gseaRes.surnag.all.GOCC, all_pathw, poi_surnag_stats,
                                out_filename = "GSEA_Output/POI_Genes_Surnageant_in_CellulCompartment_Terms.xlsx",
                                write_down = T )


```



```{r }

sessionInfo()

```
